---
title: "Introduction to eriplots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eriplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 3,
  fig.dpi = 192,
  fig.align = "center",
  out.width = 480
)
```

```{r setup}
library(ggplot2)
library(eriplots)
```

*eriplots* is an R package that produces nice-looking figures for
analysis, presentations, and reports using *ggplot2*. The package's
contributions can be organised into three categories:

1.  Overall style adjustments
2.  Helpers for saving raster (e.g., PNG) and vector (e.g., PDF) figures
3.  Elder Research color palettes

## Overall style adjustments

*eriplots* provides several styling adjustments for users:

-   A clean, minimal theme to reduce clutter
-   Theme "profiles" that tailor settings for different media
    (presentation, document, etc.)
-   Proper minus signs for continuous scales with negative values

### Opt-in vs. default

These style adjustments can be applied in two ways. First, we can use
the `eriplot()` function to create a plot with the various settings
applied automatically:

```{r}
# Make example per-manufacturer data
plotdata <- split(mpg, ~manufacturer)
plotdata <- sapply(plotdata, \(x) mean(x$cty) - mean(mpg$cty))
plotdata <- data.frame(manufacturer = names(plotdata), cty = plotdata)
plotdata$manufacturer <- with(plotdata, reorder(factor(manufacturer), cty))

eriplot(plotdata, aes(cty, manufacturer, fill = cty)) +
  geom_col(color = "black", linewidth = 0.2) +
  scale_fill_viridis_c() +
  scale_x_continuous(breaks = seq(-6, 9, 3), limits = c(-6, 9)) +
  coord_capped_cart()
```

Or, we can continue to use *ggplot2* directly and add the various pieces
manually:

```{r}
ggplot(plotdata, aes(cty, manufacturer, fill = cty)) +
  geom_col(color = "black", linewidth = 0.2) +
  scale_fill_viridis_c() +
  scale_x_continuous(breaks = seq(-6, 9, 3), limits = c(-6, 9)) +
  format_minus_signs() +
  coord_capped_cart() +
  theme_eri()
```

In this way, `eriplot()` acts as a wrapper to automatically set options
from `theme_eri()` and `format_minus_signs()`.

### Customizing default options

When using `eriplot()`, three options can be set via `options()`. These
are the defaults:

-   `eriplots.eriplot.format_minus_signs = TRUE`
-   `eriplots.eriplot.theme.apply_eri_theme = TRUE`
-   `eriplots.eriplot.theme.default_profile = "none"`

### Using the ERI theme everywhere

The base ERI theme can be set as the default by calling
`ggplot2::theme_set()` as usual:

``` r
theme_set(theme_eri())
```

As demonstrated above, it can also be composed with individual plots.

## Helpers for saving figures

The *eriplots* package also includes the function `save_figures()`,
which works with ERI theme profiles to save figures at sizes that work
well for a given destination format (presentation or document). By
default, `save_figures()` will attempt to create raster (PNG) and vector
(PDF) versions of your plot at default sizes and resolutions, but these
settings can be customized as needed. (And `ggplot2::ggsave()` will
still work just fine, too.)

```{r, eval=FALSE}
# Create figure
eriplot(plotdata, aes(cty, manufacturer, fill = cty)) +
  geom_col(color = "gray10") +
  scale_fill_viridis_c() +
  theme_eri(profile = "document")

# Save example.pdf and example.png
save_figures("example.png", formats = c("pdf", "png"))
```

Raster figures are saved at 300 DPI by default for high-quality,
high-DPI applications. If [optipng](https://optipng.sourceforge.net/) is
available, e.g., via `brew install optipng`, PNG file sizes will also be
optimized automatically.

## Elder Research color palettes

The package also bundles a color palette derived from [Elder
Research](https://www.elderresearch.com)'s brand guidelines by
[R. Robison](https://www.linkedin.com/in/robert-robison/). The colors
are available directly from the `eri_colors()` function:

```{r}
# All colors
eri_colors()

# Specific colors
eri_colors("darkred", "darkred", "lightblue")

# Specific colors with specified names
eri_colors(A = "darkred", B = "lightblue")
```

A palette is also available via `eri_palette()`. One can specify a fixed
number of colors or also provide names, which is helpful when working
with `ggplot2::scale_color_manual()`, for example:

```{r}
# Fixed number of colors
eri_palette(n = 3)

# Include names
eri_palette(n = 3, names = c("A", "B", "C"))
```

### Using the palettes with ggplot2

The palettes can be integrated directly via, e.g., `scale_fill_manual()`
for discrete sequences:

```{r}
eriplot(plotdata, aes(cty, manufacturer, fill = manufacturer)) +
  geom_col(color = "gray10") +
  scale_fill_manual(guide = "none", values = eri_palette(n = 15, recycle = TRUE))
```

Or `scale_fill_gradient()`, et al., for continuous sequences:

```{r}
eriplot(plotdata, aes(cty, manufacturer, fill = cty)) +
  geom_col(color = "gray10") +
  scale_fill_gradient2(
    low = eri_colors("darkred"),
    high = eri_colors("darkblue")
  )
```
